Challenges and Solutions of Developing and Implementing a Novel Desktop-as-a-Service

### 2024-04-29 paper.tex
Eliminate TODO's in usability section

\subsection{Usability}

An essential feature of a DaaS platform is its strong user-friendliness, which eliminates the complexity of the different operating systems used inside virtual machines and containers, makes the underlying hardware transparent, and simplifies the necessary interactions as much as possible.

Since users already have an operating system on their clients, and the motivation for using a DaaS is the simple use of various applications, it makes no sense for most users to have additional desktops available in the browser. The focus should be on the applications. For this reason, when developing a new DaaS solution, the main goal should be to reduce the graphical output to the limits of the respective application. The concept uses a separate browser tag for each application whose content exclusively represents the application.

Developing the automatic adaptation of the output to the limits of the respective application window is challenging. Only a few RDP and VNC server implementations offer to export the graphical output of only individual windows or processes. For Windows operating systems, the proprietary built-in RDP server and the free software TightVNC offer the export of full desktops and single windows. For Linux operating system deployments, the free software projects xrdp and x11vnc enable all desired features. Table ~\ref{tab:DaaS_Services_Overview} includes an overview of the mentioned service implementations for RDP and VNC protocols and their relevant characteristics. Further implementations like TigerVNC and protocols like NX NoMachine exist but do not seem useful for designing and implementing a novel DaaS because they do not meet all feature requirements.~\cite{OJCC_2023v8i1n01_Baun}

\begin{table}
	\centering
	\caption{Services used by our DaaS for Remote Access to individual Applications}
	\begin{tabular}{|l|c|c|c|c|}
		\hline
		\textbf{Service}        & \textbf{Protocol} & \textbf{Single Window} & \textbf{Operating} & \textbf{Software} \\
		\textbf{implementation} &                   & \textbf{Mode}          & \textbf{System}    & \textbf{License}  \\
		\hline
		TightVNC                & VNC               & supported              & Windows            & GPL2              \\
		\hline
		Windows RDP server      & RDP               & supported              & Windows            & proprietary       \\
		\hline
		x11vnc                  & VNC               & supported              & Linux              & GPL2              \\
		\hline
		Xrdp                    & RDP               & supported              & Linux              & Apache 2.0        \\
		\hline
	\end{tabular}
	\label{tab:DaaS_Services_Overview}
\end{table}

The automatic adaptation of browser tabs content size, is done by.... TOTO: Johannes´...

TODO: Ideen und Konzepte zur Trennung von User und Admin-Ansicht

TODO: Herausforderungen, Entwicklung, Implementierung, Lessons-learned

TODO Johannes: Erlebnisse der letzten 3-4 Wochen aufschreiben

>>>>

\subsection{Usability}
An essential feature of a DaaS platform is its strong user-friendliness,
which eliminates the complexity of the different operating systems
used inside virtual machines and containers,
makes the underlying hardware transparent,
and simplifies the necessary interactions as much as possible.

Since users already have an operating system on their clients,
and the motivation for using a DaaS is the simple use of various applications,
it makes no sense for most users to have additional desktops available in the browser.
The focus should be on the applications.
For this reason, when developing a new DaaS solution,
the main goal should be to reduce the graphical output
to the limits of the respective application.
The concept uses a separate browser tab for each application
whose content exclusively represents the application.

Developing the automatic adaptation of the output
to the limits of the respective application window is challenging.
Only a few RDP and VNC server implementations offer to export the graphical output
of only individual windows or processes.
For Windows operating systems, the proprietary built-in RDP server
and the free software TightVNC offer the export of full desktops and single windows.

For Linux operating system deployments,
the free software projects xrdp and x11vnc enable all desired features
and single-app functionality can be easily obtained
by using a custom window manager such as for example qtile
~\footnote{\url{https://qtile.org/}}.

Table~\ref{tab:DaaS_Services_Overview} includes an overview
of the mentioned service implementations for RDP and VNC protocols
and their relevant characteristics.
Further implementations like TigerVNC and protocols like NX NoMachine exist
but do not seem useful for designing and implementing a novel DaaS because
they do not meet all feature requirements.~\cite{OJCC_2023v8i1n01_Baun}

\begin{table}
	\centering
	\caption{Services used by our DaaS for Remote Access to individual Applications}
	\begin{tabular}{|l|c|c|c|c|}
		\hline
		\textbf{Service}        & \textbf{Protocol} & \textbf{Single Window} & \textbf{Operating} & \textbf{Software} \\
		\textbf{implementation} &                   & \textbf{Mode}          & \textbf{System}    & \textbf{License}  \\
		\hline
		TightVNC                & VNC               & supported              & Windows            & GPL2              \\
		\hline
		Windows RDP server      & RDP               & supported              & Windows            & proprietary       \\
		\hline
		x11vnc                  & VNC               & not supported          & Linux              & GPL2              \\
		\hline
		Xrdp                    & RDP               & supported              & Linux              & Apache 2.0        \\
		\hline
	\end{tabular}
	\label{tab:DaaS_Services_Overview}
\end{table}

The automatic adaptation of viewer size and screen resolution,
considering underlying protocols and systems,
faced various limitations, briefly described below.

One challenge was imposed by Guacamole's current lack of support
for dynamic desktop resizing with VNC,
leading to misaligned layouts and a poor user experience.
To address this until such features are available in Guacamole
\footnote{\url{https://issues.apache.org/jira/browse/GUACAMOLE-1196}},
we implemented a separate mechanism using a websocket proxy
and direct interaction with each instance.

Additionally, protocol-specific behaviors may vary
when hosted on different platforms or infrastructures.
For instance, we observed connection terminations when resizing Linux-based containers,
contrasting with the behavior observed in equivalent Linux-based virtual machines.
To overcome such limitations, we created a separate layer around the Guacamole client,
adapting to each infrastructure, platform, or protocol-specific configuration.

Another challenge was managing RDP sessions which are typically tied to actual users,
unlike VNC sessions which are typically authenticated against
a server-maintained user registry.
Utilizing the websocket proxy, the backend could pass credentials
to the server without revealing them to the user.

Essentially, Guacamole with its protocol implementations and additional features like audio and printing,
provided the fundamental base for our core capabilities.

Finally mentioned, certain instances might not be fully accessible during runtime
due to hardware malfunctions, inappropriate system configurations for simply during system updates.
Therefore, as a failover mechanism additional logic had to be added
in order to provide reasonable debug views and recovery mechanisms,
e.g.~to recreate instances from a specific snapshot or configuration state.

A crucial lesson learned was therefore to realize that
while individual problems can mostly be solved
using established paradigms, services or libraries with only slight modifications,
the real challenge for a full-fledged DaaS system lies in it's abilities
to combining such isolated solutions appropriately
and for each specific infrastructure and platform configuration.
